Week 2 Assignment
==============================

### Load Parameters and libraries

We will first load the paremeters to read in the file.

```{r Load Parameters}
fileName <- "activity.csv"

library(ggplot2)
library(dplyr)
```

### Read in the file and create a codebook

Read file into a dataframe.

```{r}

df <- read.csv(fileName, header = T)
str(df)

```

### Preprocessing

Before analyzing the dataset, we will run the following pre-processing steps.

1. Convert 'data' variable into date format
2. convert 'interval' variable to factor


Convert variables to their corresponding classes.

```{r}
df$date <- as.Date(df$date, "%Y-%m-%d")

str(df)
```

### What is the mean total number of steps taken per day?

#### Histogram of the total number of steps taken each day.

```{r Histogram, warning=F, fig.align= 'center'}
grouped <- group_by(df, date)
grouped <- summarise_all(grouped, sum)
ggplot(grouped, aes(x = steps)) + geom_histogram(bins=30) + ggtitle("Total Number of Steps per day") +
    ylab("Counts") +
    xlab("Number of steps (daily)")

```

#### Mean and Median total number of steps taken per day

```{r echo=F}

print(paste0("Average number of steps per day: ",round(mean(grouped$steps, na.rm=T),0)))

print(paste0("Median number of steps per day: ",round(median(grouped$steps, na.rm=T),0)))

```

```{r Export Histogram, echo =F}
png("./instructions_fig/Histogram.png")
grouped <- group_by(df, date)
grouped <- summarise_all(grouped, sum)
ggplot(grouped, aes(x = steps)) + geom_histogram(bins=30) + ggtitle("Total Number of Steps per day") +
    ylab("Counts") +
    xlab("Number of steps (daily)")

dev.off()

```

